Refactoring the \texttt{nls()} function in R

### 2022-07-22 RefactoringNLS/RefactoringNLS.Rmd
minor edit to termination criteria
edits to recognize developments July 2022

An important consideration is the  maintainability of
the code base that provides the functionality `nls()` claims to offer.
Our work suggests that the
existing code makes maintenance and improvement very difficult,
with legacy applications and examples blocking some important updates.
Discussion of these matters is relevant to improving R generally, as
well as its nonlinear estimation tools.

@@

- handles formulas that include R functions
- allows data to be subset
- permits parameters to be indexed over a set of related data
- produces measures of variability (i.e., standard error estimates) for the estimated parameters
- has related profiling capabilities for exploring the likelihood surface as parameters are changed
- links to a number of pre-coded ("selfStart") models

@@

This amounts to admitting R cannot solve well-posed problems unless data is polluted
with errors.

@@

The single convergence criterion of `nls()` leaves out some possibilities that could
be useful for some problems. The package `nlsr` (@nlsr-manual) already offers both 
the safeguarded relative offset test (**roffset**) as well as a **small sum of 
squares** test (**smallsstest**) that compares the latest evaluated sum of squared
(weighted) residuals to `e4` times the initial sum of squares, where
`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.

@@

Let us consider an 'singular gradient' example 
*https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values*.
`nlsr::nlxb()` displays the singular values of the Jacobian, so the `nls()` error message is correct,
but unhelpful in obtaining a solution.

@@

The Jacobian is essentially singular as shown by its singular
values. Note that these are **displayed** in a single column in
the output to provide a compact layout, but the values do **NOT** correspond to 
the individual parameters in 
whose row they appear; they are a property of the whole problem.

@@

The Jacobian used by `nlsr::nlxb()` is computed from analytic expressions, while that
of `nls()` (and, by extension, `minpack.lm::nlsLM`) mostly use a numerical approximation. 
This can lead to different behaviour:

@@

Note that the selfStart models in the base R `./src/library/stats/R/zzModels.R` file
provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines each
model, and these Jacobians may be the analytic forms. The `nls()` function, after computing the
"right hand side" or `rhs` of the residual, then checks to see if the "gradient" attribute is
defined, and, if not, uses `numericDeriv` to compute a Jacobian into that attribute. This code
is within the `nlsModel()` or `nlsModel.pliner()` functions. The use of analytic Jacobians
almost certainly contributes to the good performance of `nls()` on selfStart models.

@@

## Feature: sources of data

@@

Nonlinear estimation algorithms are almost all iterative and need a set of starting
parameters. `nls()` offers a special class of modeling functions called **selfStart** 
models. There are a number of these in base R (`./src/library/stats/R/zzModels.R`)
and others in R packages such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as 
the now-archived package `NRAIA`. Unfortunately, some **selfStart** codes
entangle the calculation of starting values for parameters
with the particulars of the `nls()` code. Though there is a `getInitial()` function,
this is not easy to use to simply compute the initial parameter estimates, in part 
because it may call `nls()`. Such circular references are, in our view, dangerous.

@@

>>>>

Important considerations are the usability and  maintainability of
the code base that provides the functionality `nls()` claims to offer.
Our work suggests that the
existing code makes maintenance and improvement very difficult,
with legacy applications and examples blocking some important updates.
Discussion of these matters is relevant to improving R generally, as
well as its nonlinear estimation tools.

@@

- handles formulas that include R functions, even ones which call calculations in
  other programming languages
- allows data to be weighted or subset
- can estimate bound constrained parameters
- provides a mechanism for handling partially linear models
- permits parameters to be indexed over a set of related data
- produces measures of variability (i.e., standard error estimates) for the estimated parameters
- has related profiling capabilities for exploring the likelihood surface as parameters are changed
- links to a number of pre-coded ("selfStart") models

@@

This amounted to admitting R cannot solve well-posed problems unless data is polluted
with errors.

@@

The convergence criterion above, the principal termination control of `nls()`, leaves out 
some possibilities that could
be useful for some problems. The package `nlsr` (@nlsr-manual) already offers both 
the safeguarded relative offset test (**roffset**) as well as a **small sum of 
squares** test (**smallsstest**) that compares the latest evaluated sum of squared
(weighted) residuals to `e4` times the initial sum of squares, where
`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.

@@

Let us consider a 'singular gradient' example 
*https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values*.
`nlsr::nlxb()` displays the singular values of the Jacobian, which confirm that the `nls()` error message is correct,
but this is not helpful in obtaining a solution.

@@

The Jacobian is essentially singular as shown by its singular
values. Note that these are **displayed** by package `nlsr` in a single column in
the output to provide a compact layout, but the values do **NOT** correspond to 
the individual parameters in 
whose row they appear; they are a property of the whole problem.

@@

As far as we can understand the logic in `nls()`, the Jacobian computation during
parameter estimation is carried out entirely within the called C-language program,
and the R code `function numericDeriv()`, part of `./src/library/stats/R/nls.R`
in the R distribution source code, is used to provide Jacobian information in 
the `nlsModel()` and `nlsModel.plinear()` functions.

The Jacobian used by `nlsr::nlxb()` is, by default, computed from analytic expressions, while that
of `nls()` mostly uses a numerical approximation. Some selfStart models do provide
"gradient" information. `minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
version of `nlsModel()`, but it appears to use an internal approximate Jacobian code from the 
original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
can lead to different behaviour, though in our experience, one has to look for the
differences, as they are generally minor.

@@

Note that the selfStart models in the base R `./src/library/stats/R/zzModels.R` file may
provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines each
model, and these Jacobians may be the analytic forms. The `nls()` function, after computing the
"right hand side" or `rhs` of the residual, then checks to see if the "gradient" attribute is
defined, and, if not, uses `numericDeriv` to compute a Jacobian into that attribute. This code
is within the `nlsModel()` or `nlsModel.plinear()` functions. The use of analytic Jacobians
almost certainly contributes to the good performance of `nls()` on selfStart models.

## Feature: Weights on observations

?? does resid return weighted or unweighted residuals

?? do we want unweighted (raw) residuals? or are predictions / fits just as good?

@@

## Feature: Sources of data

@@

Nonlinear estimation algorithms are almost all iterative and need a set of starting
parameters. `nls()` offers a special class of modeling functions called **selfStart** 
models. There are a number of these in base R (`./src/library/stats/R/zzModels.R`)
and others in R packages such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as 
the now-archived package `NRAIA`. Unfortunately, some **selfStart** codes
entangle the calculation of starting values for parameters
with the particulars of the `nls()` code. Though there is a `getInitial()` function,
this is not easy to use to simply compute the initial parameter estimates outside of
`nls()`, in part 
because it may call that function. Such circular references are, in our view, dangerous.
Moreover, we believe that it would be helpful to have selfStart models that allow
users to explicitly provide the starting parameters.

@@

We also note that the introduction of `scaleOffset` in R 4.1.1 to deal with the 
convergence test for small residual problems now requires that the `getInitial()`
function have dot-arguments (`...`) in its argument list. This illustrates the
entanglement of many features in `nls()`.

### 2022-07-22 RefactoringNLS/RefactoringNLS.tex
edits to recognize developments July 2022

An important consideration is the maintainability of the code base that
provides the functionality \texttt{nls()} claims to offer. Our work
suggests that the existing code makes maintenance and improvement very
difficult, with legacy applications and examples blocking some important
updates. Discussion of these matters is relevant to improving R
generally, as well as its nonlinear estimation tools.

@@

\begin{itemize}
\tightlist
\item
  handles formulas that include R functions
\item
  allows data to be subset
\item
  permits parameters to be indexed over a set of related data
\item
  produces measures of variability (i.e., standard error estimates) for
  the estimated parameters
\item
  has related profiling capabilities for exploring the likelihood
  surface as parameters are changed
\item
  links to a number of pre-coded (``selfStart'') models
\end{itemize}

@@

The single convergence criterion of \texttt{nls()} leaves out some
possibilities that could be useful for some problems. The package
\texttt{nlsr} (John C. Nash and Murdoch
(\protect\hyperlink{ref-nlsr-manual}{2019})) already offers both the
safeguarded relative offset test (\textbf{roffset}) as well as a
\textbf{small sum of squares} test (\textbf{smallsstest}) that compares
the latest evaluated sum of squared (weighted) residuals to \texttt{e4}
times the initial sum of squares, where
\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is
approximately 2.43e-55.

@@

Let us consider an `singular gradient' example
\emph{\url{https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values}}.
\texttt{nlsr::nlxb()} displays the singular values of the Jacobian, so
the \texttt{nls()} error message is correct, but unhelpful in obtaining
a solution.

@@

\begin{verbatim}
## nlsr object: x 
## residual sumsquares =  0  on  11 observations
##     after  1    Jacobian and  1 function evaluations
##   name            coeff          SE       tstat      pval      gradient    JSingval   
## a                     -3             0       -Inf          0           0       26.49  
## b                      5             0        Inf          0           0       9.615  
## c                    0.5             0        Inf          0           0       2.466  
## r                    0.7             0        Inf          0           0      0.1098  
## m                      1             0        Inf          0           0   1.949e-09
\end{verbatim}

@@

The Jacobian is essentially singular as shown by its singular values.
Note that these are \textbf{displayed} in a single column in the output
to provide a compact layout, but the values do \textbf{NOT} correspond
to the individual parameters in whose row they appear; they are a
property of the whole problem.

@@

The Jacobian used by \texttt{nlsr::nlxb()} is computed from analytic
expressions, while that of \texttt{nls()} (and, by extension,
\texttt{minpack.lm::nlsLM}) mostly use a numerical approximation. This
can lead to different behaviour:

@@

Note that the selfStart models in the base R
\texttt{./src/library/stats/R/zzModels.R} file provide the Jacobian in
the ``gradient'' attribute of the ``one-sided'' formula that defines
each model, and these Jacobians may be the analytic forms. The
\texttt{nls()} function, after computing the ``right hand side'' or
\texttt{rhs} of the residual, then checks to see if the ``gradient''
attribute is defined, and, if not, uses \texttt{numericDeriv} to compute
a Jacobian into that attribute. This code is within the
\texttt{nlsModel()} or \texttt{nlsModel.pliner()} functions. The use of
analytic Jacobians almost certainly contributes to the good performance
of \texttt{nls()} on selfStart models.

@@

\hypertarget{feature-sources-of-data}{%
\subsection{Feature: sources of data}\label{feature-sources-of-data}}

@@

Nonlinear estimation algorithms are almost all iterative and need a set
of starting parameters. \texttt{nls()} offers a special class of
modeling functions called \textbf{selfStart} models. There are a number
of these in base R (\texttt{./src/library/stats/R/zzModels.R}) and
others in R packages such as CRAN package \texttt{nlraa} (Miguez
(\protect\hyperlink{ref-MiguezNLRAA2021}{2021})), as well as the
now-archived package \texttt{NRAIA}. Unfortunately, some
\textbf{selfStart} codes entangle the calculation of starting values for
parameters with the particulars of the \texttt{nls()} code. Though there
is a \texttt{getInitial()} function, this is not easy to use to simply
compute the initial parameter estimates, in part because it may call
\texttt{nls()}. Such circular references are, in our view, dangerous.

@@

@@

\begin{verbatim}
## nlsr object: x 
## residual sumsquares =  9.4726  on  12 observations
##     after  16    Jacobian and  17 function evaluations
##   name            coeff          SE       tstat      pval      gradient    JSingval   
## b1                     2U           NA         NA         NA           0       23.35  
## b2               4.43325            NA         NA         NA   1.583e-10           0  
## b3                     3U           NA         NA         NA           0           0
\end{verbatim}

>>>>

Important considerations are the usability and maintainability of the
code base that provides the functionality \texttt{nls()} claims to
offer. Our work suggests that the existing code makes maintenance and
improvement very difficult, with legacy applications and examples
blocking some important updates. Discussion of these matters is relevant
to improving R generally, as well as its nonlinear estimation tools.

@@

\begin{itemize}
\tightlist
\item
  handles formulas that include R functions, even ones which call
  calculations in other programming languages
\item
  allows data to be weighted or subset
\item
  can estimate bound constrained parameters
\item
  provides a mechanism for handling partially linear models
\item
  permits parameters to be indexed over a set of related data
\item
  produces measures of variability (i.e., standard error estimates) for
  the estimated parameters
\item
  has related profiling capabilities for exploring the likelihood
  surface as parameters are changed
\item
  links to a number of pre-coded (``selfStart'') models
\end{itemize}

@@

The convergence criterion above, the principal termination control of
\texttt{nls()}, leaves out some possibilities that could be useful for
some problems. The package \texttt{nlsr} (John C. Nash and Murdoch
(\protect\hyperlink{ref-nlsr-manual}{2019})) already offers both the
safeguarded relative offset test (\textbf{roffset}) as well as a
\textbf{small sum of squares} test (\textbf{smallsstest}) that compares
the latest evaluated sum of squared (weighted) residuals to \texttt{e4}
times the initial sum of squares, where
\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is
approximately 2.43e-55.

@@

Let us consider a `singular gradient' example
\emph{\url{https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values}}.
\texttt{nlsr::nlxb()} displays the singular values of the Jacobian,
which confirm that the \texttt{nls()} error message is correct, but this
is not helpful in obtaining a solution.

@@

\begin{verbatim}
## nlsr object: x 
## residual sumsquares =  0  on  11 observations
##     after  1    Jacobian and  1 function evaluations
##   name            coeff          SE       tstat      pval      gradient    JSingval   
## a                     -3            NA         NA         NA           0       26.49  
## b                      5            NA         NA         NA           0       9.615  
## c                    0.5            NA         NA         NA           0       2.466  
## r                    0.7            NA         NA         NA           0      0.1098  
## m                      1            NA         NA         NA           0   1.398e-16
\end{verbatim}

@@

The Jacobian is essentially singular as shown by its singular values.
Note that these are \textbf{displayed} by package \texttt{nlsr} in a
single column in the output to provide a compact layout, but the values
do \textbf{NOT} correspond to the individual parameters in whose row
they appear; they are a property of the whole problem.

@@

As far as we can understand the logic in \texttt{nls()}, the Jacobian
computation during parameter estimation is carried out entirely within
the called C-language program, and the R code
\texttt{function\ numericDeriv()}, part of
\texttt{./src/library/stats/R/nls.R} in the R distribution source code,
is used to provide Jacobian information in the \texttt{nlsModel()} and
\texttt{nlsModel.plinear()} functions.

The Jacobian used by \texttt{nlsr::nlxb()} is, by default, computed from
analytic expressions, while that of \texttt{nls()} mostly uses a
numerical approximation. Some selfStart models do provide ``gradient''
information. \texttt{minpack.lm::nlsLM()} invokes
\texttt{numericDeriv()} in its local version of \texttt{nlsModel()}, but
it appears to use an internal approximate Jacobian code from the
original Fortran \texttt{minpack} code, namely, \texttt{lmdif.f}. Such
differences in approach can lead to different behaviour, though in our
experience, one has to look for the differences, as they are generally
minor.

@@

Note that the selfStart models in the base R
\texttt{./src/library/stats/R/zzModels.R} file may provide the Jacobian
in the ``gradient'' attribute of the ``one-sided'' formula that defines
each model, and these Jacobians may be the analytic forms. The
\texttt{nls()} function, after computing the ``right hand side'' or
\texttt{rhs} of the residual, then checks to see if the ``gradient''
attribute is defined, and, if not, uses \texttt{numericDeriv} to compute
a Jacobian into that attribute. This code is within the
\texttt{nlsModel()} or \texttt{nlsModel.plinear()} functions. The use of
analytic Jacobians almost certainly contributes to the good performance
of \texttt{nls()} on selfStart models.

\hypertarget{feature-weights-on-observations}{%
\subsection{Feature: Weights on
observations}\label{feature-weights-on-observations}}

?? does resid return weighted or unweighted residuals

?? do we want unweighted (raw) residuals? or are predictions / fits just
as good?

@@

\hypertarget{feature-sources-of-data}{%
\subsection{Feature: Sources of data}\label{feature-sources-of-data}}

@@

Nonlinear estimation algorithms are almost all iterative and need a set
of starting parameters. \texttt{nls()} offers a special class of
modeling functions called \textbf{selfStart} models. There are a number
of these in base R (\texttt{./src/library/stats/R/zzModels.R}) and
others in R packages such as CRAN package \texttt{nlraa} (Miguez
(\protect\hyperlink{ref-MiguezNLRAA2021}{2021})), as well as the
now-archived package \texttt{NRAIA}. Unfortunately, some
\textbf{selfStart} codes entangle the calculation of starting values for
parameters with the particulars of the \texttt{nls()} code. Though there
is a \texttt{getInitial()} function, this is not easy to use to simply
compute the initial parameter estimates outside of \texttt{nls()}, in
part because it may call that function. Such circular references are, in
our view, dangerous. Moreover, we believe that it would be helpful to
have selfStart models that allow users to explicitly provide the
starting parameters.

@@

We also note that the introduction of \texttt{scaleOffset} in R 4.1.1 to
deal with the convergence test for small residual problems now requires
that the \texttt{getInitial()} function have dot-arguments
(\texttt{...}) in its argument list. This illustrates the entanglement
of many features in \texttt{nls()}.

@@

\begin{verbatim}
## nlsr object: x 
## residual sumsquares =  9.4726  on  12 observations
##     after  16    Jacobian and  17 function evaluations
##   name            coeff          SE       tstat      pval      gradient    JSingval   
## b1                     2U           NA         NA         NA           0       23.35  
## b2               4.43325            NA         NA         NA   1.832e-10           0  
## b3                     3U           NA         NA         NA           0           0
\end{verbatim}
